Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-06 ms.tex
addtional comments by MWF; initial integration of modifier model in ms

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem*{theorem}{Theorem}
\newtheorem*{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof

@@

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find a general and simple result:
if individuals with below-average fitness increase the rate at which variation is generated, then the population mean fitness increases.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary (possibly infinite) set of types $A_1, A_2, \ldots$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots)$ is a stochastic vector: $\sum_k{f_k} = 1$);
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_j {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_k{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).
We focus on a modifier that has no direct effect on fitness and whose sole function is to determine the rate of transition out of the different types ($C_k$).
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.

In the following section we present three particular examples of this model for the modification of mutation, migration, and learning.

\paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}

@@

\paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype.
A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_k{f_k}=1$.
Then after selection and mutation $\vec{f'}$ in the next generation is given by
\begin{equation}
f_0' = (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1},
\end{equation}
and
\begin{multline}
f_k' = (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
\delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\label{eq:mutation_model_2}
\end{multline}
for all $k=1,2,\ldots$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k \ge 0}{w_k f_k}$ is the population mean fitness.

@@

% Modifier model
\paragraph*{Modifier model.}
We now consider the case of two modifier alleles, $m$ and $M$, in complete linkage to the types $A_k$.
These modifiers potentially induce different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ is $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
the rest of the model parameters are the same as in Eq.~\ref{eq:model};
and the frequencies in the next generation, $f'$ and $g'$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

@@

\begin{mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:sign_theorem}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if and only if the fitness of type $k$ is below the stable population mean fitness.
\end{mfp}

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$).

@@

\begin{rvp}
In the notation of the theorem,
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if and only if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{rvp}

@@

We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{eq:sign_theorem_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:sign_theorem}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if and only if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

@@

We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
But could such a modifier increase in frequency when initially rare in the population?
To answer this question we will analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eq.~\ref{eq:modifier_model}.

The equilibrium of Eq.~\ref{eq:modifier_model} when $M$ is missing from the population ($g_k=0$ for all $k$) is (modified from \ref{eq:model_equilibrium})
\begin{equation}\label{eq:modifier_model_equilibrium}
\bar{w} \cdot \begin{pmatrix}v \\ 0\end{pmatrix} = 
\Big(\vec{I} - \vec{C} + \vec{M} \vec{C}\Big)\vec{D} \begin{pmatrix}v \\ 0\end{pmatrix},
\end{equation}
where $\vec{I}$ is a $2n$-by-$2n$ identity matrix, and
\begin{equation}
%\begin{aligned}
\vec{C} =  \begin{pmatrix}C & 0 \\ 0 & \tilde{C} \end{pmatrix}, \quad
\vec{M} =  \begin{pmatrix}M & 0 \\ 0 & M \end{pmatrix}, \quad
\vec{D} =  \begin{pmatrix}D & 0 \\ 0 & D \end{pmatrix}.
%\end{aligned}
\end{equation}

The stability of modifier $m$ to invasion by modifier $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $L_{ex}$ of modifier $M$, which is in turn determined by the Jacobian of the system in Eq.~\ref{eq:modifier_model_equilibrium}
\begin{equation}
J = \begin{pmatrix}L_{in} & 0 \\ 0 & L_{ex} \end{pmatrix}.
\end{equation}
Note the zero block matrices due to complete linkage between the modifier and the types $A_k$.

$L_{ex}$ can be written 
\begin{equation}\label{eq:Lex}
L_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness without modifier $M$ (see~\nameref{sec:AppA}).

\begin{theorem}[Evolution of increased genetic variation]
The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $L_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}
In addition, the leading eigenvalue is $\lambda=1$ when transition rates are equal $\tilde{C}_k=C_k$ for all $k$.

Therefore, an invading modifier $M$ with transition rates slightly different from the resident modifier $m$ can successful invade the population ($\lambda>1$ if $M$ increases the probability of transition from types with below-average fitness.
\end{theorem}

\begin{proof}
First, consider the case the $\tilde{C_k} = C_k$ for all $k$.
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* L_{ex} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
then the leading eigenvalue of $L_{ex}$ is $\lambda=1$.

Now, applying Eq.~\ref{eq:sign_theorem} on Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
\end{equation}
and evaluating this expression at $\lambda=1$ (the case for $\tilde{C}_k = C_k$) gives us
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
sign(\bar{w}^* - D_k),
\end{equation}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

@@

% Appendix C: Fisher's reproductive value
\subsection*{Appendix C}\label{sec:AppC}
\begin{frv}
Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
\end{frv}

>>>>

% Yoav & Lee commands
\newcommand*{\tr}{^\intercal}
\let\vec\mathbf
\newcommand{\matrx}[1]{{\left[ \stackrel{}{#1}\right]}}
\newcommand{\diag}[1]{\mbox{diag}\matrx{#1}}
\newcommand{\goesto}{\rightarrow}
\newcommand{\dspfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2} }
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}
%\newtheorem*{mfp}{Theorem: Mean Fitness Principle}
%\newtheorem*{rvp}{Corollary: Reproductive Value Principle}
%\newtheorem*{frv}{Fisher's reproductive values}
\renewcommand\qedsymbol{} % no square at end of proof
\newcommand{\cl}{\mathbf{L}}
\newcommand{\cj}{\mathbf{J}}
\newcommand{\ci}{\mathbf{I}}

@@

% Abstract
\begin{abstract}
Generation of variation may be detrimental in well-adapted populations evolving under constant selection.
In a constant environment, genetic modifiers that reduce the rate at which variation is generated by processes such as mutation and migration, succeed.
However, departures from this \emph{reduction principle} have been demonstrated.
Here we analyze a general model of evolution under constant selection where the rate at which variation is generated can depend on the fitness of the individual.
We find a general and clear result:
if a modifier allele increases the rate at which below-average individuals generate variation, then it will increase in frequency and increase the  population mean fitness.
This principle applies to phenomena such as stress-induced mutagenesis, condition-dependent dispersal and outcrossing,
and even to the proverb: \emph{"Necessity is the mother of genetic invention"}.
\end{abstract}

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The change in the frequencies of these types is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f_k$ is the frequency of type $A_k$ ($f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector: $\sum_{k=1}^n{f_k} = 1$);
$C_k$ is the probability that an individual of type $A_k$ will transition to another type ($C$ is a diagonal non-negative matrix with diagonal elements $C_k$);
$M_{i,j}$ is the transition probability from type $A_j$ to type $A_i$ given that an individual of type $A_j$ is transitioning ($M$ is a column-stochastic matrix: $\sum_{j=1}^n {M_{i,j}} = 1$ for all \emph{j}); 
$D_k$ is the relative fitness of type $A_k$ ($D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type); 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

Transition between types is determined by a combination of two effects:
(i) the probability to transition \emph{out} of type $A_k$ is determined by $C_k$;
(ii) given a transition out of type $A_k$, the distribution of the destination types $A_j$ is given by $M_{j,k}$ (note the index order).

We focus on a neutral modifier locus in complete linkage to the types $A_k$, with no direct effect on fitness, and whose sole function is to determine the rate of transition out of the different types ($C_k$).
Importantly, this modifier can determine different rates for different types; that is, $C_j \ne C_i$ for some $i,j$. The case $C_j = C_j$ for all $i,j$ is covered by the \emph{reduction principle}, see~\citet{Altenberg2017}.

% Modifier model
\paragraph*{Modifier model.}
Considering the case of two modifier alleles, $m$ and $M$, inducing different transition probabilities out of types $A_k$, with transition matrices $C=diag[C_1, \ldots, C_n]$ and $\tilde{C}=diag[\tilde{C}_1, \ldots, \tilde{C}_n]$.
The frequencies of type $A_k$ linked to modifier $m$ or $M$ are $f_k$ and $g_k$, respectively, such that $\sum_k{f_k+g_k}=1$;
$\bar{w}$ now ensures that $\sum_{k}{f_k'+g_k'}=1$;
the rest of the model parameters are the same as in Eq.~\ref{eq:model};
and the frequencies in the next generation, $f'$ and $g'$, are given by
\begin{equation}
\begin{cases}
\bar{w} f' &= (I-C+MC)D f \\
\bar{w} g' &= (I-\tilde{C}+M\tilde{C})D g
\end{cases}.
\label{eq:modifier_model}
\end{equation}
Note that Eq.~\ref{eq:model} is a special case for $g_k=0$ for all $k$.
Here, $\bar{w}=\sum_{k}{(f_k+g_k)D_k}$ is the mean fitness of the entire population (see~\nameref{sec:AppA}).

In the following section we present three particular examples of the model presented in Eq.~\ref{eq:model} for the modification of mutation and migration.

% Mutation model 1
\paragraph*{Mutation modification model 1.}\label{sec:mut_model_1}

@@

% Mutation model 2
\paragraph*{Mutation modification model 2.}\label{sec:mut_model_2}

Here, we consider a large population of haploids in which individuals with genotype $A_k$ are characterized by the number $k$ of deleterious or mutant alleles in their genotype, where $0 \le k \le n$.
A modifier locus then determines the probability $m_k$ that a mutation occurs in individuals with $k$ mutant alleles.
In this case, we assume a low mutation rate $m_k$ so that two mutations are unlikely to occur in the same individual in one generation.
When a mutation occurs it is \emph{deleterious} with probability $\delta$, generating a mutant allele and converting the individual from $A_k$ to $A_{k+1}$, or it is beneficial with probability $\beta$, converting the individual from $A_k$ to $A_{k-1}$.
Note that such beneficial mutations can be either compensatory or back-mutations, and that mutations are neutral with probability $1-\delta-\beta$.
This model has been analyzed by~\citet{Ram2012}.

Let the frequency of $A_k$ in the present generation be $f_k$ such that $f_k \ge 0$ and $\sum_{k=0}^{n}{f_k}=1$.
Then after selection and mutation $f'$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= (1 - \delta m_0) \frac{w_0}{\bar{w}} f_0 + \beta m_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= (1 - \beta m_n) \frac{w_n}{\bar{w}} f_n + \delta m_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= (1 - m_k(\delta+\beta)) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta m_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta m_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$, where $w_k$ is the fitness of an individual with $k$ mutant alleles and $\bar{w} = \sum_{k=0}^{n}{w_k f_k}$ is the population mean fitness.

@@

@@

\begin{lemma}[Mean fitness principle]
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{eq:theorem}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{eq:sign_theorem}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

Therefore increased transition from type $k$ will increase the stable population mean fitness if the fitness of type $k$ is below the stable population mean fitness.
\end{lemma}

@@

\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial C_k} = 
v_k D_k (u\tr[M]_k - u_k) = \\
D_k (\sum_i{u_i M_{i,k} v_k} - u_k v_k).
\end{aligned}
\end{equation}
Here, $u_k$ can be regarded as the \emph{reproductive value}~\citep[pg.~27]{Fisher1930} of type $k$ (see~\nameref{sec:AppC}), which gives the relative contribution of type $k$ to the long-term population.
Consequently, $u_k v_k$ is the \emph{ancestor frequency} \citep{Hermisson2002} of type $k$, namely the fraction of the equilibrium population descended from type $k$.
The sum $\sum_i{u_i M_{i,k} v_k}$ can be similarly interpreted as the fraction of the equilibrium population descended from individuals that transitioned from type $k$ to another type (via the $k$ column of the transition matrix $M$), conditioned on transition occuring.

@@

\begin{corollary}[Reproductive value principle]
In the notation of the theorem,
\begin{equation}
sign \frac{\partial \bar{w}^*}{\partial C_k} = 
sign (u\tr [M]_k - u_k),
\end{equation}
and therefore increased transition out of type $k$ will increase the stable population mean fitness if the fraction of the population descended from type $k$ is expected to increase due to a transition to another type. 
\end{corollary}

@@

We turn our attention to a modifier that concurrently controls the transition out of a subset of types, $K$.
For example, \citet[Appendix~B]{Ram2012} considered individuals that are grouped by the number of their accumulated deleterious mutations, $k$, and a modifier that regulates the mutation rates in individuals with at least $\pi$ deleterious mutations.
Similarly, if the transition rates are controlled by a threshold rule 
\begin{equation}
C_k = \begin{cases}
\tau, & k \in K \\
1, & k \not\in K
\end{cases},
\end{equation}
where $\tau > 0$, then the sensitivity of the stable population mean fitness to changes in the
rate of transition from types $K$ can be written 
\begin{equation}
\begin{aligned}
\frac{\partial \bar{w}^*}{\partial \tau} = 
u\tr \Big( \sum_{k \in K}{\frac{\partial (I - C + MC)D}{\partial \tau}} \Big) v = \\
\frac{1}{\tau} \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)},
\end{aligned}
\end{equation}
which can be simplified to
\begin{multline}\label{eq:sign_theorem_threshold}
sign \frac{\partial \bar{w}^*}{\partial \tau} = 
sign \sum_{k \in K}{u_k v_k (\bar{w}^* - D_k)} = \\ 
sign \Big(\bar{w}^* - \frac{\sum_{k \in K}{u_k v_k D_k}}{\sum_{k \in K}{u_k v_k}}\Big), 
\end{multline}
a generalization of Eq.~\ref{eq:sign_theorem}.
Therefore, increased transition from types $K$ will increase the stable population mean fitness if the average fitness of individuals descended from types $K$ is below the stable population mean fitness.

@@

We determined a condition for a modifier that controls that transition rate $C_k$ from type $A_k$ to increase the stable population mean fitness.
Could such a modifier increase in frequency when initially rare in the population?
To answer this question we analyze the stability of modifier $m$ with transition rates $C_k$ to the invasion of modifier $M$ with rates $\tilde{C}_k$, as described in Eqs.~\ref{eq:modifier_model}.

The equilibrium of Eqs.~\ref{eq:modifier_model} when modifier allele $M$ is absent from the population is $(v, 0)$, where $v$ is given in eq.~\ref{eq:model_equilibrium} and $g_k=0$ for all $k$. 
The stability of modifier allele $m$ to invasion by allele $M$ is determined by the leading eigenvalue $\lambda$ of the external stability matrix $\cl_{ex}$ of the equilibrium $(v,0)$, which, in turn, is determined by the Jacobian $\cj$ of Eq.~\ref{eq:modifier_model} evaluated at the equilibrium $(v,0)$, where
\begin{equation}
\cj = \begin{pmatrix}\cl_{in} & 0 \\ 0 & \cl_{ex} \end{pmatrix}.
\end{equation}
The zero block matrices are due to complete linkage between the modifier and the types $A_k$.

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\bar{w}^*,
\end{equation}
where $\bar{w}^*$ is the stable population mean fitness in the absence of modifier allele $M$ (see~\nameref{sec:AppA}).

\begin{theorem}[Evolution of increased genetic variation]
The sign of the derivative of the leading eigenvalue $\lambda$ of the external stability matrix of $\cl_{ex}$ of modifier $M$, when evaluated at $\tilde{C}_k=C_k$, is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}
In addition, the leading eigenvalue is $\lambda=1$ when transition rates are equal $\tilde{C}_k=C_k$ for all $k$.

Therefore, an initially rate modifier allele $M$ with transition rates slightly different from the resident modifier $m$ can successfully invade the population ($\lambda>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

\begin{proof}
First, consider the case the $\tilde{C_k} = C_k$ for all $k$.
Substituting $\tilde{C}=C$ in Eq.~\ref{eq:Lex} and multiplying both sides by $\bar{w}^*$,
\begin{equation}
\bar{w}^* L_{ex} = (I-C+MC) D,
\end{equation}
and since $\bar{w}^*$ is the leading eigenvalue of the RHS (\nameref{sec:AppA}),
the leading eigenvalue of $\cl_{ex}$ is $\lambda=1$.

Now, applying Eq.~\ref{eq:sign_theorem} to Eq.~\ref{eq:Lex}, the sign of the derivative of $\lambda$ with respect to $\tilde{C_k}$ is
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k} =
sign\Big(\lambda - \frac{D_k}{\bar{w}^*}\Big),
\end{equation}
and evaluating this expression at $\lambda=1$ (the case for $\tilde{C}_k = C_k$) gives us
\begin{equation}
sign \frac{\partial \lambda}{\partial \tilde{C}_k}\bigg\rvert_{\lambda=1} = 
sign(\bar{w}^* - D_k),
\end{equation}
since $\bar{w}^*>0$.
This completes the proof.
\end{proof}

@@

% Appendix C: Fisher's reproductive value
\subsection*{Appendix C}\label{sec:AppC}
\begin{remark}[Fisher's reproductive value]
Let $M$ be a stochastic matrix and $D$ be a diagonal matrix with positive diagonal elements. 
The elements of the left \emph{Perron} eigenvector $u$ of the matrix $MD$ can be regarded as \emph{Fisher's reproductive values}~\citep[pg.~27]{Fisher1930}
\end{remark}
